---
title: "5-Minute Quickstart"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{5-Minute Quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Installation

First, install metasalmon from GitHub:

```{r}
# Install from GitHub (recommended)
install.packages("remotes")
remotes::install_github("dfo-pacific-science/metasalmon")
```

## What You'll Learn

By the end of this guide, you'll be able to:

- Turn your salmon data spreadsheet into a shareable "data package"
- Create a data dictionary that explains what each column means
- Share data that colleagues can immediately understand

## What You'll Need

- R installed (version 4.4 or higher)
- Your salmon data as a CSV file (or use our example data)
- About 5 minutes

## Video Version

Prefer video? [Watch the walkthrough](https://youtu.be/B0Zqac49zng?si=VmOjbfMDMd2xW9fH)

---

## Step 1: Load Your Data

First, let's load the metasalmon package and some example data. We'll use a sample of NuSEDS escapement data for Fraser River coho that comes with the package.

```{r}
library(metasalmon)

# Load the example data included with the package
data_path <- system.file("extdata", "nuseds-fraser-coho-sample.csv",
                          package = "metasalmon")
df <- readr::read_csv(data_path, show_col_types = FALSE)

# Take a look at what we have
head(df)
```

**What you see**: A table with columns like `POP_ID`, `SPECIES`, `ANALYSIS_YR`, `MAX_ESTIMATE`, etc.

**The problem**: What does `POP_ID` mean? What are the valid values for `SPECIES`? If you shared this CSV with a colleague, they'd have questions.

> **Using your own data?** Just replace the `data_path` line with your file path:
> ```r
> df <- readr::read_csv("path/to/your-data.csv", show_col_types = FALSE)
> ```

---

## Step 2: Generate a Data Dictionary

metasalmon can look at your data and create a starter dictionary - a table that describes each column.

```{r}
dict <- infer_dictionary(
  df,
  dataset_id = "fraser-coho-2024",
  table_id = "escapement"
)

# See what it created
print(dict)
```

**What you get**: A table with one row per column in your data, describing:

| Field | What it means |
|-------|---------------|
| `column_name` | The column name from your data |
| `column_label` | A human-readable label (you can edit this) |
| `value_type` | Is it text (`string`), a number (`integer`, `number`), a date? |
| `column_role` | Is this an ID, a measurement, a category, or a date? |

The dictionary is a starting point - metasalmon makes educated guesses, but you can (and should) review and improve the descriptions.

> **Tip**: To see all columns in the dictionary, use `View(dict)` in RStudio or `print(dict, width = Inf)`.

### Need Help Finding Standard Terms?

Not sure what the official salmon data standard term is for a column? The `suggest_semantics()` function can automatically suggest standard terminology from scientific vocabularies:

```{r}
# Get semantic suggestions for your dictionary
# For salmon-domain roles, SMN shared terms are queried first, then GCDFO as a
# distinct DFO-specific source before OLS/NVS fallback sources.
dict_suggested <- suggest_semantics(df, dict)

# View the suggestions
suggestions <- attr(dict_suggested, "semantic_suggestions")
head(suggestions)
```

This searches standard ontologies and vocabularies to find matching terms for your columns, helping you link your data to recognized scientific standards without making OLS/NVS the first stop for salmon-domain matches.

> **Want faster results?** Use the [Salmon Data Standardizer GPT](https://chatgpt.com/g/g-69375eab4f608191863e8c23313a6f9f-salmon-data-standardizer) to get AI-powered suggestions for terminology, descriptions, and metadata. Just upload your dictionary and data sample!

```{r}
# Optional: include DwC-DP export mappings alongside ontology suggestions
sem <- suggest_semantics(dict)                  # default: DwC export off
sem_with_dwc <- suggest_semantics(dict, include_dwc = TRUE)
```

DwC-DP mappings stay optional; keep SDP columns as canonical and use the DwC view only when exporting to biodiversity tooling.

### Step 2b: Decide whether missing terms belong in SMN or in a project/program vocabulary

Not every domain term should be forced into SMN. Use this as a practical review step:

```{r}
# Identify measurement columns still without SMN-style IRIs
measurement_cols <- dict$column_name[dict$column_role == "measurement"]
missing_smn <- dict$column_name[is.na(dict$term_iri) & dict$column_name %in% measurement_cols]

# Optional: run broader sources for these terms to see candidates
# Good candidates here can indicate:
# - likely SMN addition candidate (shared across programs), or
# - dataset-local/program-specific term for a custom controlled vocabulary
scope_review <- purrr::map_dfr(missing_smn, function(col) {
  q <- dict$column_label[dict$column_name == col][1]
  if (is.na(q) || q == "") q <- col

  broader <- find_terms(
    query = q,
    sources = c("gcdfo", "ols", "nvs")
  )

  if (nrow(broader) == 0) {
    return(tibble::tibble(column_name = col, candidate_label = NA_character_, candidate_iri = NA_character_, source = NA_character_))
  }

  tibble::tibble(
    column_name = col,
    candidate_label = broader$label[1],
    candidate_iri = broader$iri[1],
    source = broader$source[1],
    score = broader$score[1]
  )
})

scope_review
```

Use this review table as a **scope decision**:
- If a term is broadly used across many salmon datasets and has clear ontological intent, use the result as a **SMN ontology addition request** path.
- If it is specific to one project/program/organization workflow (e.g., local ID patterns, internal quality flags, station code conventions), keep it in a **local controlled vocabulary**.

---

## Step 3: Check Your Dictionary

Before packaging, let's make sure the dictionary is valid:

```{r}
validate_dictionary(dict)
```

**What happens**:

- **Green checkmarks** = everything looks good
- **Warnings** = optional improvements you could make
- **Errors** = things you need to fix before proceeding

If you see errors, the message will tell you what's wrong. Common fixes:

```{r}
# Example: Fix a column type that was guessed incorrectly
dict$value_type[dict$column_name == "YEAR"] <- "integer"

# Example: Add a better description
dict$column_description[dict$column_name == "POP_ID"] <-
  "Unique population identifier from the NuSEDS database"

# Validate again
validate_dictionary(dict)
```

---

## Step 4: Describe Your Dataset

Now we need to add some basic information about the dataset as a whole - who created it, what it contains, and how others can use it.

```{r}
# Dataset-level metadata (describes the overall dataset)
dataset_meta <- tibble::tibble(
  dataset_id = "fraser-coho-2024",
  title = "Fraser River Coho Escapement Data",
  description = "Sample escapement monitoring data for coho salmon in PFMA 29",
  creator = "DFO Pacific Science",
  contact_name = "Your Name",
  contact_email = "your.email@dfo-mpo.gc.ca",
  license = "Open Government License - Canada",
  temporal_start = "2001",
  temporal_end = "2024",
  spatial_extent = "PFMA 29, Fraser River watershed",
  # Optional but useful for EDH/GeoNetwork-ready export
  contact_org = "Fisheries and Oceans Canada - Pacific Region Science Branch",
  contact_position = "Fishery and Assessment Data Section",
  update_frequency = "annually",
  topic_categories = "biota;oceans",
  keywords = "coho;escapement;Fraser River",
  security_classification = "unclassified"
)

# Table-level metadata (describes this specific table)
table_meta <- tibble::tibble(
  dataset_id = "fraser-coho-2024",
  table_id = "escapement",
  file_name = "escapement.csv",
  table_label = "Escapement Data",
  description = "Coho escapement counts by population and year",
  primary_key = "POP_ID"
)
```

**What these fields mean**:

| Field | Purpose |
|-------|---------|
| `dataset_id` | A short identifier (letters, numbers, hyphens) |
| `title` | Human-readable title for the dataset |
| `description` | What the data contains |
| `creator` | Who created/collected the data |
| `contact_name/email` | Who to contact with questions |
| `license` | How others can use the data |
| `contact_org`, `contact_position`, `update_frequency`, `topic_categories`, `keywords`, `security_classification` | Optional discovery/governance fields (recommended for EDH-ready export) |
| `table_id` | Identifier for this table (matches `dict`) |
| `file_name` | What to name the output CSV file |

---

## Step 5: Create Your Data Package

Now let's bundle everything together into a shareable package:

```{r}
resources <- list(
  escapement = df
)

# Recommended for fast first-pass bootstrap
# Optional: include EDH XML in the package with the default HNAP-aware profile
pkg_path <- create_salmon_datapackage_from_data(
  resources = resources,
  path = "my-first-package",
  dataset_id = "fraser-coho-2024",
  table_id = "escapement",
  overwrite = TRUE,
  include_edh_xml = TRUE,
  edh_profile = "dfo_edh_hnap"
)

# Optional alternative: write ISO19139 profile instead
# pkg_path <- create_salmon_datapackage_from_data(
#   resources = resources,
#   path = "my-first-package-iso",
#   dataset_id = "fraser-coho-2024",
#   table_id = "escapement",
#   overwrite = TRUE,
#   include_edh_xml = TRUE,
#   edh_profile = "iso19139"
# )

# See what was created
list.files(pkg_path)
```

> **Bootstrap note:** `create_salmon_datapackage_from_data()` is for fast first-pass output, not final publication.
> Before sharing, run the full validation + explicit semantic workflow:
> 1) `validate_dictionary()` and `validate_semantics()`
> 2) `suggest_semantics()`
> 3) `apply_semantic_suggestions()`
> 4) `create_salmon_datapackage()`

For full control or staged review, use the explicit workflow in the
[Publishing Data Packages](https://dfo-pacific-science.github.io/metasalmon/articles/data-dictionary-publication.html) guide.

```{r}
# Optional explicit control (no bootstrap assumptions):
# artifacts <- infer_salmon_datapackage_artifacts(
#   resources,
#   dataset_id = "fraser-coho-2024",
#   seed_semantics = TRUE
# )
#
# validate_dictionary(artifacts$dict)
# validate_semantics(artifacts$dict)
# dict <- apply_semantic_suggestions(artifacts$dict, max_per_role = 3)
# pkg_path <- create_salmon_datapackage(
#   resources = resources,
#   dataset_meta = artifacts$dataset_meta,
#   table_meta = artifacts$table_meta,
#   dict = dict,
#   codes = artifacts$codes,
#   path = "my-first-package",
#   overwrite = TRUE
# )
```

**What you get**: A folder called `my-first-package/` containing:

| File | Purpose |
|------|---------|
| `escapement.csv` | Your data |
| `column_dictionary.csv` | What each column means |
| `datapackage.json` | Machine-readable metadata |
| `dataset.csv` | Dataset-level information |
| `tables.csv` | Table-level information |
| `metadata-edh-hnap.xml` | Optional EDH metadata export (if included in one-shot call) |

---

## Step 5b (Optional): Export EDH XML metadata

If your workflow includes DFO Enterprise Data Hub / GeoNetwork upload, the
default `edh_build_iso19139_xml()` path now writes the richer HNAP-aware EDH
XML export from `dataset_meta`. The older compact ISO 19139 shape is still
available as an explicit fallback.

```{r}
edh_hnap_xml_path <- file.path(pkg_path, "metadata-edh-hnap.xml")
edh_build_iso19139_xml(dataset_meta, output_path = edh_hnap_xml_path)

edh_iso_xml_path <- file.path(pkg_path, "metadata-iso19139.xml")
edh_build_iso19139_xml(
  dataset_meta,
  output_path = edh_iso_xml_path,
  profile = "iso19139"
)

file.exists(edh_hnap_xml_path)
file.exists(edh_iso_xml_path)
```

The default HNAP-aware path adds EDH-oriented structure like
maintenance/status, legal constraints, optional download/distribution metadata,
reference system info, bounding boxes, deterministic identifiers, and bilingual
locale scaffolding. Use `profile = "iso19139"` only when you specifically need
that smaller fallback export.

Validate and enrich either XML output against your local EDH profile before
production upload.

---

## Step 6: Share It!

Your data package is ready. You can:

- **Email the folder** to a colleague (zip it first)
- **Upload to a data repository** like Zenodo or CIOOS
- **Archive it** for your future self
- **Include it** in a research compendium

When someone opens your package, they'll find not just data, but complete documentation explaining what every column means.

---

## Reading a Package Back

Later, you (or a colleague) can load the package back into R:

```{r}
# Read the package
pkg <- read_salmon_datapackage(pkg_path)

# What's inside?
names(pkg)

# Access the components
pkg$dataset      # Dataset metadata
pkg$tables       # Table metadata
pkg$dictionary   # Column descriptions
pkg$resources    # Your actual data

# Get your data back as a tibble
head(pkg$resources$escapement)
```

---

## What's Next?

You've created your first Salmon Data Package! Here are some ways to go deeper:

- **[Using AI to Document Your Data](gpt-collaboration.html)** - Use the Salmon Data Standardizer GPT to get AI-powered suggestions for terminology, descriptions, and metadata
- **[Publishing Data Packages](data-dictionary-publication.html)** - More control over metadata and publishing
- **[Linking to Standard Vocabularies](reusing-standards-salmon-data-terms.html)** - Connect your data to scientific standards
- **[Accessing Data from GitHub](github-csv-access.html)** - Read CSVs from private repositories
- **[Glossary of Terms](glossary.html)** - Definitions of technical terms
- **[FAQ](faq.html)** - Common questions and troubleshooting

---

## Troubleshooting

### "validate_dictionary() shows errors"

This usually means a column type was guessed incorrectly. Check the error message and fix:

```{r}
# See what types are valid
# string, integer, number, boolean, date, datetime

# Fix a specific column
dict$value_type[dict$column_name == "PROBLEM_COLUMN"] <- "string"
```

### "Column not found in dictionary"

Make sure your `table_id` in `table_meta` matches the `table_id` you used in `infer_dictionary()`.

### "I don't understand what a field means"

See the [Glossary](glossary.html) for plain-English definitions of terms like "column_role", "value_type", etc.

### "I want to add better descriptions"

Edit the dictionary directly before creating the package:

```{r}
# View and edit in RStudio
View(dict)

# Or edit programmatically
dict$column_description[dict$column_name == "MAX_ESTIMATE"] <-
  "Maximum escapement estimate for the population in a given year"
dict$column_label[dict$column_name == "MAX_ESTIMATE"] <-
  "Maximum Estimate"
```

### Still stuck?

- [Report a bug](https://github.com/dfo-pacific-science/metasalmon/issues)
- [Ask a question](https://github.com/dfo-pacific-science/metasalmon/issues)
