---
title: "5-Minute Quickstart"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{5-Minute Quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Installation

```{r}
install.packages("remotes")
remotes::install_github("dfo-pacific-science/metasalmon")
```

## One-shot Workflow

Load the built-in NuSEDS Fraser Coho sample and create a review-ready Salmon Data Package in one call.

```{r}
library(metasalmon)

sample_path <- system.file("extdata", "nuseds-fraser-coho-sample.csv", package = "metasalmon")
fraser_coho <- readr::read_csv(sample_path, show_col_types = FALSE)

pkg_path <- create_sdp(
  fraser_coho,
  dataset_id = "fraser-coho-2024",
  table_id = "escapement",
  overwrite = FALSE
)

pkg_path
list.files(pkg_path, recursive = TRUE)
```

If `path` is omitted, `create_sdp()` writes to your working directory using a default folder name like `fraser-coho-2024-sdp`. In interactive use it can also mention when a newer `metasalmon` release is available; set `check_updates = FALSE` to skip that check.

If you also need the DFO Enterprise Data Hub / GeoNetwork XML, use the one-shot path:

```{r}
pkg_path <- create_sdp(
  fraser_coho,
  dataset_id = "fraser-coho-2024",
  table_id = "escapement",
  include_edh_xml = TRUE,
  overwrite = FALSE
)
```

That writes the HNAP-aware EDH XML to `metadata/metadata-edh-hnap.xml`. If you are working from an existing `dataset.csv` row instead of a one-shot package build, call `edh_build_hnap_xml()` directly.

## Pick The Right Bundled Example

`metasalmon` ships two Fraser coho example tables:

| File | Rows | Years | Best for |
| --- | ---: | --- | --- |
| `nuseds-fraser-coho-sample.csv` | 30 | 1996-2024 | Fastest walkthroughs and smallest demo payload |
| `nuseds-fraser-coho-2023-2024.csv` | 173 | 2023-2024 | More realistic package creation and testing |

```{r}
fuller_path <- system.file("extdata", "nuseds-fraser-coho-2023-2024.csv", package = "metasalmon")
provenance_path <- system.file("extdata", "example-data-README.md", package = "metasalmon")
```

The fuller table is an official Open Government Canada NuSEDS slice from the
Fraser and BC Interior workbook, filtered to `SPECIES == "Coho"` and
`ANALYSIS_YR %in% c(2023, 2024)`. It keeps a compact column subset and uses
`NATURAL_ADULT_SPAWNERS` because `NATURAL_SPAWNERS_TOTAL` is blank for this
two-year official slice. See `example-data-README.md` for provenance, row
counts, and the Open Government Licence - Canada note.

## Review In Excel

Open the output folder and review these files:

- `README-review.txt`
- `semantic_suggestions.csv` (when suggestions were found)
- `metadata/dataset.csv`
- `metadata/tables.csv`
- `metadata/column_dictionary.csv`
- `metadata/codes.csv` (when present)
- `data/*.csv` resource files

`create_sdp()` seeds semantic suggestions by default and auto-fills the top-ranked **column-level** suggestions into missing dictionary fields (`term_iri`, `property_iri`, `entity_iri`, `unit_iri`, etc.). It also auto-fills top table-level observation-unit suggestions into missing `tables.csv$observation_unit_iri` values (and labels when blank). It does not overwrite existing non-empty semantic values. Code-level suggestions default to factor and low-cardinality character source columns; use `semantic_code_scope = "all"` if you want broader code-level seeding.

The inferred metadata includes `MISSING DESCRIPTION:` and `MISSING METADATA:` placeholders for required fields so the package is immediately reviewable in Excel. Replace those placeholders before publishing. The `metadata/*.csv` files are the canonical package metadata; `datapackage.json` is a derived export for interoperability.

## How To Decide If `term_iri` Is Correct

Use plain-language checks for each measurement column:

1. Does the suggested label describe exactly what the column measures?
2. Does the definition match your intent (not just a similar word)?
3. Is the scope right (for example species-level vs population-level)?
4. Is the unit consistent with your values and `unit_iri`?

Keep the IRI only when all checks pass.

Replace it when the term is close but not exact.

Remove it (leave blank) when no candidate is reliable yet.

When the top auto-applied suggestion is wrong, use `semantic_suggestions.csv` to pick a better alternative and copy that IRI into `metadata/column_dictionary.csv`.

## Finalize

After Excel edits, save the metadata back to CSV, share the whole folder (or a zip of the whole folder) when you hand it to someone else, then run validation again before publishing:

```{r}
pkg <- read_salmon_datapackage(pkg_path)
validate_dictionary(pkg$dictionary)
validate_semantics(pkg$dictionary)
```

For a staged, fully explicit workflow (manual artifact inference and controlled semantic merges), see the publication guide:

- [Publishing Data Packages](https://dfo-pacific-science.github.io/metasalmon/articles/data-dictionary-publication.html)
